A Variant of G\"{o}del's Ontological Proof \\ in a Natural Deduction Calculus

### 2015-09-05 Papers/InProgress/NaturalDeduction/GodProof-ND.tex
Derivation of Brouwer Reduction Principle added
Merge Annika's contributios and include Scott's proof

\title{A Variant of G\"{o}del's Ontological Proof \\
in a Natural Deduction Calculus \\
(Draft)}

@@

\institute{
  Helsinki, Finland\\
  \email{annika.siders@helsinki.fi}
  \and 
  Vienna, Austria \\
  \email{bruno@logic.at}
}

@@

\noindent
``There is a scientific (exact) philosophy and theology,
which deals with concepts of the highest abstractness; and this is also most highly fruitful for science. [\ldots] Religions are, for the most part, bad; but religion is not.'' - Kurt G\"{o}del

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considered a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence.

A major critique of this argument is that we do not know whether the concept greatest conceivable being in fact designates anything or if it is inconsistent, like a round square \citep{fitting}[p.134].  
%The question is if the argument is sound. 
As Bertrand Russell pointed out, the definition of greatest allows us to define properties, like having boots, which the greatest being then also must have \citep{citation needed}.

Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. All that we can say is that if God exists, then God necessarily exists \citep{citation needed}.

St. Anselm's argument was elaborated further by Descartes and Leibniz. Leibniz identified that establishing the possible existence of God is a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This implies that it is possible to have all perfections at once and therefore the existence of a greatest being with all these properties is possible.

@@

G\"odel's notion of positive property and Leibniz's notion of perfection are not exactly the same \citep{Adams}. On a technical level, the main distinction seems to be that G\"odel's positive properties are not just atomic properties, like Leibniz's perfections, but can also consist of complex combinations of atomic properties \citep{fitting}[p.139.] (TODO: Check this claim). In particular, one of G\"odel's axioms states that the conjunction of any set of positive properties is positive. And from this axiom, it follows immediately that the property of being God-like is positive. While this step is intuitively and informally clear, it is not easily formalizable in a typical logical calculus, because it requires inferring that being god-like is a (possibly infinite) conjuntion of positive properties, while it has only been defined as the property holding for individuals who have all positive properties. This interplay between universal quantification (in the definition of a God-like being) and conjunction (in G\"odel's mentioned axiom) is a technical difficulty that probably explains why, starting with Scott \cite{scott}, to whom G\"odel confided his manuscript, this axiom of G\"odel's has been replaced by another one that simply assumes the positivity of the property of being God-like.

The main criticism against G\"odel's ontological argument is the \emph{modal collapse}, an undesirable consequence of the stipulated axioms. Most recent works \cite{todo} on the ontological argument have focused on proposing modifications of the argument that would not entail a modal collapse. This is discussed in more detail in Section \ref{sec:collapse}.

@@

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus (as defined in Section \ref{sec:calculus}). A natural deduction style \citep{gentzen,prawitz} was chosen mainly for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Finally, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

The main contribution of the paper, however, is the presentation of new proofs (also in natural deduction style) of the lemmas, theorems and corollaries in G\"odel's manuscript. In contrast to Scott's proofs \cite{Scott}, the proofs presented here are simpler and shorter, as discussed in Section \ref{sec:conclusion}.

@@

Adding the modal rules results in a calculus that is suitable for the basic normal modal logic \textbf{K}.
In other words, {\NDK} is sound and complete relative to {\ND} extended with axiom K ($\nec(A\imp B)\imp (\nec A\imp \nec B)$) and the necessitation rule (which establishes that $\nec A$ is a theorem if $A$ is a theorem).

@@

\begin{small}
\begin{prooftree}
\AXC{$\nec(A\imp B)^2 $}\RightLabel{$\nec_E$}
\UIC{$\omega: A\imp B$}
      \AXC{$\nec A ^1 $}\RightLabel{$\nec_E$}
      \UIC{$\omega: A$} \RightLabel{$\imp_E$}
   \BIC{$ \omega: B$} \RightLabel{$\nec_I$}
   \UIC{$ \nec B$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec B$} \RightLabel{$\imp_I^2$}
   \UIC{$\nec(A\imp B)\imp (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

@@

ToDo: show example of unsound derivation.

@@

\section{A distribution principle and axiom B}

To prove G\"odel's ontological proof in natural deduction we will prove that a distribution principle is derivable and state Brouwer's reduction theorem.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \dashedLine\RightLabel{2}
\UIC{$\nec (A\imp B)$}\RightLabel{$\nec_E$}
\UIC{$\omega: A\imp B$}
\AXC{$ $} \dashedLine\RightLabel{1}
\UIC{$\pos A$}\RightLabel{$\pos_E$}
\UIC{$\omega: A$}\RightLabel{$\imp_E$}
\BIC{$\omega: B$} \RightLabel{$\pos_I$}
\UIC{$\pos B$}\RightLabel{$\imp_I^1$}
\UIC{$\pos A\imp \pos B$}\RightLabel{$\imp_I^2$}
\UIC{$\nec (A\imp B)\imp(\pos A\imp \pos B)$}
\end{prooftree}
\end{small}

@@

Brouwer's reduction theorem is a version of axiom B for modal logic and a property that enforces symmetry as a geometric property for the frame of a Kripke sematics for modal logic. \citep{negri}
%See the following for a discussion about axiom B: http://plato.stanford.edu/entries/logic-modal/

\section{Scott's Proof in Natural Deduction}

\setcounter{axiom}{0}
\setcounter{lemma}{0}
\setcounter{theorem}{0}
\setcounter{corollary}{0}

@@

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if God's existence is possible, then it is necessary and a proof that God's existence is in fact possible.

@@

\section{New Proof}
\setcounter{axiom}{0}
\setcounter{lemma}{0}
\setcounter{theorem}{0}
\setcounter{corollary}{0}
\subsection{Possibly, God Exists}
\begin{axiom}
\label{A1}
Either a property or its negation is positive, but not both:
Axiom schema, for all $\varphi$:
$$
P(\neg \varphi) \biimp \neg P(\varphi)
$$
\end{axiom}
\begin{axiom}
\label{A2}
A property necessarily implied by a positive property is positive:
Axiom schema, for all $\varphi$ and $\psi$: 
$$
(P(\varphi) \wedge \nec \all x.[\varphi(x) \imp \psi(x)]) \imp P(\psi)
$$
\end{axiom}
\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
Axiom schema, for all $\varphi$:
$$
P(\varphi) \imp \pos \ex x.\varphi(x)
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}

\AXC{$[P(\rho)]^3$} 
\AXC{$[\nec \neg \ex x.\rho(x) ]^2$} \RightLabel{$\nec_E$}
\UIC{$\neg \ex x.\rho(x) $} 
\AXC{$[\rho(x)]^1$} \RightLabel{$\ex_I$}
\UIC{$\ex x.\rho(x) $} \RightLabel{$\imp_E$}
\BIC{$ \bot $}\RightLabel{$\imp_I, 1$}
\UIC{$ \neg \rho(x) $} \RightLabel{$\imp_I$}
\UIC{$ \rho(x) \imp \neg \rho(x) $} \RightLabel{$\all_I$}
\UIC{$ \all x.(\rho(x) \imp \neg \rho(x)) $}  \RightLabel{$\nec_I$}
\UIC{$ \nec \all x.(\rho(x) \imp \neg \rho(x)) $}\RightLabel{$\wedge_I$}
\BIC{$ P(\rho) \wedge \nec \all x.[\rho(x) \imp \neg \rho(x)]$}
\AXC{Axiom \ref{A2} for $\rho$ and $\neg\rho$} \dashedLine
\UIC{$(P(\rho) \wedge \nec \all x.[\rho(x) \imp \neg \rho(x)]) \imp P(\neg \rho)$} \RightLabel{$\imp_E$}
\BIC{$P(\neg \rho)$} \RightLabel{$\imp_E$}
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \dashedLine
\UIC{$P(\neg \rho)$} \RightLabel{$\imp_E$}
\AXC{Half of axiom \ref{A1}} \dashedLine
\UIC{$ P(\neg \rho) \imp \neg P(\rho) $}  \RightLabel{$\imp_E$}
\BIC{$\neg P(\rho) $} 
\AXC{$[P(\rho)]^3$} 
\BIC{$\bot$} \RightLabel{$\imp_I, 2$}
\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \dashedLine \RightLabel{definition of $\pos$}
\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I, 3$}
\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $}

@@

\end{proof}
\begin{definition}
\label{D1}
A \emph{God-like} being possesses all positive properties:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
$$
\end{definition}

\begin{axiom}
\label{A3}
The property of being God-like is positive:
$$
P(G)
$$
\end{axiom}
\begin{corollary}
\label{C1}
Possibly, God exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof} \hfill
\begin{prooftree}
\AXC{Axiom \ref{A3}} \dashedLine
\UIC{$P(G)$}
\AXC{Theorem \ref{T1} for G} \dashedLine
\UIC{$ P(G) \imp \pos \ex x.G(x) $} \RightLabel{$\imp_E$}
\BIC{$\pos \ex x. G(x)$}
\end{prooftree}
\end{proof}

>>>>

\title{A Variant of G\"{o}del's Ontological Proof \\
in a Natural Deduction Calculus}

@@

\institute{
  Helsinki, Finland\\
  \email{annika.siders@helsinki.fi}
  \and 
  Vienna, Austria ; Canberra, Australia \\
  \email{bruno.wp@gmail.com}
}

@@

\noindent
\begin{footnotesize}
\begin{center}
``There is a scientific (exact) philosophy and theology,
which deals with concepts of the highest abstractness; and this is also most highly fruitful for science. [\ldots] \\Religions are, for the most part, bad; but religion is not.'' \\
- Kurt G\"{o}del \citep{Wang1996}[p. 316]
\end{center}
\end{footnotesize}

%Begin with a focus on more recent stuff such as Gödel, not Anselm and so on. Focus on the improved level of formalization of the argument through Leibniz/Descartes/Kant and so on. 
%Think about how much can be claimed about Russell, Descartes, Kant and so on... 
%Russell eliminated or expained as a valuable contribution to the formal proof. 
%check the Solmon paper to se about proofs in KB. Other ontological proofs. 
%Relate introduction to the beginning quote
%spell check
%How to use the words God, God-like, god-like and greatest conceivable being. 
%Theorem reference missing in the proof of modal collapse
%Write the section on axiom B and the strength of the modal system that is necessary.

@@

Ontological arguments for the existence of God can be traced back at least to St. Anselm (1033-1109). His argument considers a greatest conceivable being, who must exist, because if it did not have the property of existence, then we could conceive of a greater being that, apart from the other properties, also has the property of existence.

A major critique of this argument is that we do not know whether the concept of greatest conceivable being in fact designates anything or if it is inconsistent, like a round square \citep{fitting}[p.134].  
%The question is if the argument is sound. 
As Bertrand Russell pointed out, the definition of greatest conceivable being allows us to define properties, like having boots, which the being then also must have \citep{citation needed}.

Kant argued against the ontological argument on the basis that existence is not an analytic property \citep{kant}. This means that existence cannot be contained in the definition of a concept, because it is generally synthetic. The strongest claim that can be proven by the ontological argument is a conditional claim of necessary existence: if God exists, then God necessarily exists \citep{citation needed}.

St. Anselm's argument was further elaborated by Descartes and Leibniz. Leibniz identified the possible existence of God as a critical missing step in St. Anselm's argument. To fill this gap, he argued that the properties of God, the perfections, are compatible. This means that it is possible to satisfy all perfections at once which implies that 
the existence of a greatest conceivable being with all these properties is possible.

@@

G\"odel's notion of positive property and Leibniz's notion of perfection differ \citep{Adams}. The main formal distinction seems to be that G\"odel's positive properties are not just atomic properties, like Leibniz's perfections, but can also consist of complex combinations of atomic properties \citep{fitting}[p.139.] (TODO: Check this claim). In particular, one of G\"odel's axioms states that any conjunction of positive properties is itself positive. From this axiom, it is immediately deduced that the property of being God-like is positive. While this step is intuitively and informally clear, it is not easily formalizable in a standard logical calculus, because it requires inferring that being God-like is a (possibly infinite) conjunction of positive properties, from the universal definition of God-likeness as the property satisfied by individuals who have all positive properties. This interplay between universal quantification (in the definition of a God-like being) and infinite conjunctions (in G\"odel's axiom for positive properties) is a technical difficulty that probably explains why, starting with Scott \cite{Scott}, to whom G\"odel confided his manuscript, this axiom of G\"odel's has been replaced by another that simply assumes the positivity of the property of being God-like.

The main criticism against G\"odel's ontological argument is an undesirable consequence of the stipulated axioms, called \emph{modal collapse}. Many recent works \cite{todo} on the ontological argument have focused on proposing modifications of the argument that would not entail a modal collapse. This is discussed in greater detail in Section \ref{sec:collapse}, but possible solutions are not implemented in this paper.

@@

The aim of this paper is to present two detailed formal proofs of G\"odel's ontological argument \cite{Goedel} in a natural deduction calculus (as defined in Section \ref{sec:calculus}). The natural deduction style \citep{gentzen,prawitz} was chosen for three reasons. Firstly, presentations of G\"odel's proof are typically either informal or formalized in other styles of calculi (e.g. Fitting's tableaux \cite{fitting} or Sobel's sentential modal calculus \cite{sobel2}). Therefore, a formalization in natural deduction is a valuable complement to the existing presentations. Secondly, it makes the ontological proof accessible to people who are familiarized with a natural deduction style. Thirdly, as natural deduction is the style used by proof assistants such as Coq \cite{coq} and Isabelle \cite{isabelle}, the natural deduction formalizations have been easily checked step-by-step in Coq \cite{coqformalizations}.

The first contribution of this paper is a detailed formalization of Scott's version \cite{Scott} of G\"odel's ontological argument \cite{Goedel} (as shown in Section \ref{sec:scott}) in a natural deduction calculus. The second contribution of the paper is a new proof (also in natural deduction style) of the lemmas, theorems and corollaries in G\"odel's manuscript. In contrast to Scott's proofs \cite{Scott}, the proofs presented here are simpler and shorter, as discussed in Section \ref{sec:conclusion}.

@@

Adding the modal rules results in a calculus that is suitable for the basic modal logic \textbf{K}.
In other words, {\NDK} is sound and complete relative to {\ND} extended with axiom K ($\nec(A\imp B)\imp (\nec A\imp \nec B)$) and the necessitation rule (which establishes that $\nec A$ is a theorem if $A$ is a theorem).

@@

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{2}
\UIC{$\nec(A\imp B)$}\RightLabel{$\nec_E$}
\UIC{$\omega: A\imp B$}
      \AXC{$ $}\RightLabel{1}
      \UIC{$\nec A $}\RightLabel{$\nec_E$}
      \UIC{$\omega: A$} \RightLabel{$\imp_E$}
   \BIC{$ \omega: B$} \RightLabel{$\nec_I$}
   \UIC{$ \nec B$} \RightLabel{$\imp_I^1$}
   \UIC{$\nec A\imp \nec B$} \RightLabel{$\imp_I^2$}
   \UIC{$\nec(A\imp B)\imp (\nec A\imp \nec B)$}
\end{prooftree}
\end{small}

@@

\begin{prooftree}
  \AXC{$ $} \RightLabel{1}
  \UIC{$\nec \psi$} \RightLabel{$\nec_E$}
        \UIC{$\omega: \psi$} \RightLabel{$\pos_I$}
        \UIC{$\pos \psi$} \RightLabel{$\imp_I^1$}
    \UIC{$\nec \psi \imp \pos \psi$} \RightLabel{$\forall_I$}
    \UIC{$\all \psi. (\nec \psi \imp \pos \psi)$}
\end{prooftree}

This example of an unsound derivation is prevented by the eigen-box condition because the box labelled by $\omega$ is not accessed by any strong inference.

@@

\section{Some Derivable Modal Principles}
\label{sec:AxiomB}.

As mentioned in \citep{and90}[Footnote 5], there are sources that discuss what modal principles are needed for the ontological argument. The standard system for the argument is S5. Through an analysis of the proofs weaker systems, in particular KB, can be shown to be sufficient for versions of the ontological argument \citep{sobel2}[p. 152]. The proof presented in Section \ref{sec:newproof} is such a version and explicitly shows a single use of Brouwer Reduction Principle, which, in turn, is provable from axiom B. All other proof steps are formalizable in a basic second-order modal logic K.

In standard first-order modal logics the modal axioms correspond to geometric conditions on the frames of Kripke models \citep{negri}. The modal axiom {\bf T} corresponds to reflexivity of the frame and {\bf B} corresponds to symmetry of the frame. The frames of the modal logic {\bf S5} are in addition transitive.

%G\"odel mentions in the notes which are saved of his proof that the modal system required for proving his theorem is a second order version of the modal system {\bf S5}. In \citep{and90} it is conjectured that the principles of proof could be restricted to the weaker modal logic {\bf KB}. The modal logic {\bf KB} consists of the basic modal system {\bf K} and the modal axioms {\bf T} and {\bf B}.

ToDo: Sobel Anderson on KB: Sobel page 152
and
Anderson's footnote 5 on KB %Nathan Salmon: On the Logic of What Might Have Been,"

\subsection{A modal distribution principle in K}

To prove G\"odel's ontological proof in natural deduction we use a modal distribution principle derivable in K and  Brouwer's reduction theorem derivable from the modal Axiom B.

@@

\begin{small}
\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\nec (A\imp B)$}\RightLabel{$\nec_E$}
\UIC{$\omega: A\imp B$}
\AXC{$ $} \RightLabel{1}
\UIC{$\pos A$}\RightLabel{$\pos_E$}
\UIC{$\omega: A$}\RightLabel{$\imp_E$}
\BIC{$\omega: B$} \RightLabel{$\pos_I$}
\UIC{$\pos B$}\RightLabel{$\imp_I^1$}
\UIC{$\pos A\imp \pos B$}\RightLabel{$\imp_I^2$}
\UIC{$\nec (A\imp B)\imp(\pos A\imp \pos B)$}
\end{prooftree}
\end{small}

@@

\subsection{Axiom B and Brouwer's Reduction Principle}

Brouwer's reduction theorem is a version of axiom B for modal logic and a property that enforces symmetry as a geometric property for the frame of a Kripke sematics.

\begin{lemma}\label{BRP}
The Brouwer Reduction Principle $\pos\nec A\imp A$ is derivable from axiom B, $A\imp \nec \pos A$. 
\end{lemma}

\begin{small}
\begin{prooftree}
\AXC{$ $}\RightLabel{$2$}
\UIC{$\neg\nec \neg \nec A$}
\AXC{$ $}\RightLabel{$ 1$}
\UIC{$\neg A$}
\AXC{Axiom B}\dashedLine
\UIC{$\neg A\imp \nec\neg\nec\neg\neg A$}\RightLabel{$\imp_E$}
\BIC{$\nec\neg\nec\neg\neg A$}\doubleLine
\UIC{$\nec\neg\nec A$}
\BIC{$(\neg\nec \neg \nec A)\&(\nec\neg \nec A)$}\RightLabel{$\neg_I^1$}
\UIC{$\neg\neg A$} \RightLabel{$\neg\neg_E$}
\UIC{$A$} \RightLabel{$\imp_I^2$}
\UIC{$\neg\nec \neg \nec A\imp A$}
\end{prooftree}
\end{small}

\subsection{S5 and its Iteration Principle}

ToDo: Is this principle really commonly called ``S5 Iteration Principle''?

\begin{lemma}
What is possibly necessary is necessary:
$$
\pos \nec A \imp \nec A
$$
\end{lemma}

\begin{proof}

\begin{prooftree}
\AXC{ToDo}
\end{prooftree}

\section{Scott's Proof in Natural Deduction}\label{sec:scott}

\setcounter{axiom}{0}
\setcounter{lemma}{0}
\setcounter{theorem}{0}
\setcounter{corollary}{0}
\setcounter{definition}{0}

@@

The acceptance of the correctness of the ontological argument by G\"odel's work boils down to the intuitive correctness of the axioms and definitions and the belief in the soundness of the deductive system. The formal argument of G\"odel is based on Leibniz proof, which in turn is based on Descartes proof. These proofs have two parts; a proof that if the exemplification of the god-likeness property is possible, then it is necessary and a proof that the exemplification is in fact possible.

@@

\begin{theorem}
\label{T1}
Positive properties are possibly exemplified:
$$
P(\varphi) \imp \pos \ex x.\varphi(x)
$$
\end{theorem}
\begin{proof} \hfill

\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$}
		\AXC{Reflexivity} \dashedLine
		\UIC{$\omega: \gamma = \gamma$} \RightLabel{$\imp_I$}
		\UIC{$\omega: \rho{\gamma} \imp \gamma = \gamma$} \RightLabel{$\all_I$}
		\UIC{$\omega: \all x. \rho(x) \imp x = x$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp x = x) $} \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x = x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x = x) \imp P(\lambda x. x = x)$}

		\BIC{$P(\lambda x. x = x)$} 
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \RightLabel{3}
\UIC{$P(\rho)$}
	\AXC{$ $} \RightLabel{8}
	\UIC{$\omega: \rho(\beta)$}
			\AXC{$ $} \RightLabel{5}
			\UIC{$\nec \all x. \neg \rho(x)$} \RightLabel{$\nec_E$}
			\UIC{$\omega: \all x. \neg \rho(x) $} \RightLabel{$\all_E$}
			\UIC{$\omega: \neg \rho(\beta)$} \RightLabel{$\neg_E$}
		\BIC{$\omega: \bot$} \RightLabel{$\neg_I^9$}
		\UIC{$\omega: \neg (\beta = \beta)$} \dottedLine
		\UIC{$\omega: (\beta \neq \beta)$} \RightLabel{$\imp_I^8$}
		\UIC{$\omega: (\rho(\beta) \imp (\beta \neq \beta))$} \RightLabel{$\all_I$}
		\UIC{$\omega: (\all x. \rho(x) \imp (x \neq x))$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\all x. \rho(x) \imp (x \neq x))$}  \RightLabel{$\wedge_I$}
	\BIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) $} 
			\AXC{Axiom 2 for $\rho$ and $\lambda x. x \neq x$} \dashedLine
	    	\UIC{$P(\rho) \wedge \nec (\all x. \rho(x) \imp x \neq x) \imp P(\lambda x. x \neq x)$} 
		\BIC{$P(\lambda x. x \neq x)$} 
\end{prooftree}

\begin{prooftree}
		\AXC{$ $} \dashedLine
		\UIC{$P(\lambda x. x \neq x)$}    
    		\AXC{$ $} \dashedLine
			\UIC{$P(\lambda x. x = x)$} 
					\AXC{Half of Axiom 1} \dashedLine
					\UIC{$P(\lambda x. x = x) \imp \neg P(\lambda x. x \neq x)$} \RightLabel{$\imp_E$}
				\BIC{$\neg P(\lambda x. x \neq x)$} \RightLabel{$\neg_E$}
			\BIC{$\bot$}
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \RightLabel{2}
\UIC{$\box \neg \ex x. \rho(x)$} \RightLabel{$\nec_E$}
\UIC{$\omega: \neg \ex x. \rho(x)$}
        \AXC{$ $} \RightLabel{4}
        \UIC{$\rho(\alpha)$} \RightLabel{$\ex_I$}
		\UIC{$\omega: \ex x. \rho(x)$} \RightLabel{$\neg_E$}
    \BIC{$\omega: \bot$} \RightLabel{$\neg_I^4$}
	\UIC{$\omega: \neg \rho(\alpha)$}
	\UIC{$\omega: \all x. \neg \rho(x)$}
	\UIC{$\nec \all x. \neg \rho(x)$}
	        \AXC{$ $} \dashedLine
	        \UIC{$\bot$} \RightLabel{$\neg_I^5$}
    		\UIC{$\neg \nec \all x. \neg \rho(x)$} \RightLabel{$\neg_E$}
  		\BIC{$\bot$} \RightLabel{$\neg_I^2$}
  		\UIC{$ \neg \nec \neg \ex x.\rho(x)$} \doubleLine %\RightLabel{definition of $\pos$}
  		\UIC{$ \pos \ex x.\rho(x)$}  \RightLabel{$\imp_I^3$}
  		\UIC{$ P(\rho) \imp \pos \ex x.\rho(x) $}

@@

\begin{definition}
\label{D1}
A \emph{god-like} being possesses all positive properties:
$$
G(x) \biimp \forall \varphi. [P(\varphi) \to \varphi(x)]
$$
\end{definition}

\begin{axiom}
\label{A3}
The property of being god-like is positive:
$$
P(G)
$$
\end{axiom}
\begin{corollary}
\label{C1}
Possibly, God exists:
$$
\pos \ex x. G(x)
$$
\end{corollary}
\begin{proof} \hfill
\begin{prooftree}
\AXC{Axiom \ref{A3}} \dashedLine
\UIC{$P(G)$}
\AXC{Theorem \ref{T1} for G} \dashedLine
\UIC{$ P(G) \imp \pos \ex x.G(x) $} \RightLabel{$\imp_E$}
\BIC{$\pos \ex x. G(x)$}
\end{prooftree}
\end{proof}

\begin{axiom}
\label{A4}
Positive properties are necessarily positive:
$$
\all \varphi.[P(\varphi) \to \Box \; P(\varphi)]
$$
\end{axiom}

\begin{theorem}
\label{T2}
Being God-like is an essence of any God-like being:
$$
\all x. G(x) \imp \ess{G}{x}
$$
\end{theorem}
\begin{proof}

\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$G(\gamma)$}

	\AXC{$ $} \RightLabel{2}
	\UIC{$\rho(\gamma)$}

		\AXC{$ $} \RightLabel{4}
		\UIC{$\neg P(\rho)$} 
				\AXC{Axiom 1 for $\rho$} \RightLabel{$\imp_E$}
			\BIC{$P(\lambda x. \neg \rho(x))$}

					\AXC{$ $} \RightLabel{1}
					\UIC{$G(\gamma)$} \dottedLine
					\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
					\UIC{$P(\lambda x. \neg \rho(x)) \imp \neg \rho(\gamma)$} \RightLabel{$\imp_E$}
				\BIC{$\neg \rho(\gamma)$} \RightLabel{$\neg_E$}      
	   \BIC{$\bot$} \RightLabel{$\neg_I^4$}
	   \UIC{$\neg\neg P(\rho)$} \RightLabel{$\neg\neg_E$}
	   \UIC{$P(\rho)$}
	      \AXC{Axiom 4} \dashedLine
	   	  \UIC{$P(\rho) \imp \nec P(\rho)$} \RightLabel{$\imp_E$}
		 \BIC{$\nec P(\rho)$}

			\AXC{$ $} \RightLabel{5}
			\UIC{$\omega: P(\rho)$}

					\AXC{$ $} \RightLabel{6}
					\UIC{$\omega: G(\delta) $} \dottedLine
					\UIC{$\omega: \all \varphi. P(\varphi) \imp \varphi(\delta)$} \RightLabel{$\all_E$} 
					\UIC{$\omega: P(\rho) \imp \rho(\delta)$} \RightLabel{$\imp_E$}
				\BIC{$\omega: \rho(\delta) $} \RightLabel{$\imp_I^6$}
				\UIC{$\omega: G(\delta) \imp \rho(\delta) $} \RightLabel{$\all_I$}
				\UIC{$\omega: \all y. G(y) \imp \rho(y) $} \RightLabel{$\imp_I^5$}
				\UIC{$\omega: P(\rho) \imp (\all y. G(y) \imp \rho(y) )$}
				\UIC{$\nec (P(\rho) \imp (\all y. G(y) \imp \rho(y) ))$}
						\AXC{K} \RightLabel{$\imp_E$}
					\BIC{$\nec P(\rho) \imp \nec (\all y. G(y) \imp \rho(y) )$}
			\BIC{$\nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\imp_I$}
			\UIC{$\rho(\gamma) \imp \nec (\all y. G(y) \imp \rho(y) )$} \RightLabel{$\all_I$}
			\UIC{$\all q. q(\gamma) \imp \nec (\all y. G(y) \imp q(y) )$} \RightLabel{$\wedge_I$}
		\BIC{$ G(\gamma) \wedge (\all q. q(\gamma) \imp \nec (\all y. G(y) \imp q(y) ))$} \dottedLine
		\UIC{$\ess{G}{\gamma}$} \RightLabel{$ \imp_I^1$}
		\UIC{$G(\gamma) \imp \ess{G}{\gamma}$} \RightLabel{$\all_I$}
		\UIC{$\all x. G(x) \imp \ess{G}{x}$}
\end{prooftree}

\begin{definition}
\label{D3}
\emph{Necessary existence} of an individual is the necessary exemplification of all its essences:
$$
E(x) \biimp \all \varphi.[\ess{\varphi}{x} \imp \nec \ex y.\varphi(y)]
$$
\end{definition}

\begin{axiom}
\label{A5}
Necessary existence is a positive property:
$$
P(E)
$$
\end{axiom}

\begin{lemma}
\label{L1}
Exemplification of the god-likeness property implies the exemplification is necessary:
$$
\ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

\begin{proof}

\begin{prooftree}
\AXC{$ $} \RightLabel{1}
\UIC{$\ex z. G(z)$} \RightLabel{$\ex_E$}
\UIC{$G(\gamma)$}
\end{prooftree}

\begin{prooftree}
\AXC{$ $} \dashedLine
\UIC{$G(\gamma)$}
		\AXC{Theorem 2} \dashedLine 
		\UIC{$\all x. G(x) \imp \ess{G}{x}$} \RightLabel{$\all_E$}
		\UIC{$G(\gamma) \imp \ess{G}{\gamma}$}
	\BIC{$\ess{G}{\gamma}$}
				\AXC{Axiom 5} \dashedLine
				\UIC{$P(E)$}
						\AXC{$ $} \dashedLine
						\UIC{$G(\gamma)$} \dottedLine
						\UIC{$\all \varphi. P(\varphi) \imp \varphi(\gamma)$} \RightLabel{$\all_E$}
						\UIC{$P(E) \imp E(\gamma)$} 
					\BIC{$E(\gamma)$} \dottedLine
					\UIC{$\all \varphi. \ess{\varphi}{\gamma} \imp \nec \ex x. \varphi(x)$} \RightLabel{$\all_E$}
					\UIC{$\ess{G}{\gamma} \imp \nec \ex x. G(x)$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\ex z. G(z) \imp \nec \ex x. G(x)$}

\begin{lemma}
\label{L2}
If the existence of a God-like being is possible, then it is necessary:
$$
\pos \ex z. G(z) \imp \nec \ex x. G(x)
$$
\end{lemma}

\begin{proof}

\begin{footnotesize}
\begin{prooftree}

\AXC{$ $} \RightLabel{1}
\UIC{$\pos \ex x. G(x)$} 
		\AXC{Lemma 1} \dashedLine
		\UIC{$\omega: \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\nec_I$}
		\UIC{$\nec (\ex x. G(x) \imp \nec \ex x. G(x)) $}
				\AXC{Distribution Principle}
			\BIC{$\pos \ex x. G(x) \imp \pos \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
	\BIC{$\pos \nec \ex x. G(x)$}
			\AXC{S5 Iteration Principle} \dashedLine
			\UIC{$\pos \nec \ex x. G(x) \imp \nec \ex x. G(x)$}
		\BIC{$\nec \ex x. G(x)$} \RightLabel{$\imp_I^1$}
		\UIC{$\pos \ex z. G(z) \imp \nec \ex x. G(x)$}
\end{prooftree}
\end{footnotesize}

\begin{theorem}
\label{T3}
Necessarily, there exists a God-like being:
$$
\nec \ex x. G(x)
$$
\end{theorem}

\begin{proof}

\begin{prooftree}
	\AXC{Corollary 1} \dashedLine
	\UIC{$\pos \ex x. G(x)$}
			\AXC{Lamma 2} \dashedLine
			\UIC{$\pos \ex x. G(x) \imp \nec \ex x. G(x)$} \RightLabel{$\imp_E$}
		\BIC{$\nec \ex x. G(x)$}
\end{prooftree}

\section{New Proof}\label{sec:newproof}

A outline of the proof is that firstly all positive properties are proven to be possibly exemplified (Theorem \ref{T1}). Then this theorem is applied to the property of god-likeness, thus showing the possible existence of a god-like being (Corollary \ref{C1}). In Lemma \ref{L1} it is classically proven that the hypothetically assumed existence of a god-like being implies that the existence is necessary. Lastly, the necessary existence of a god-like being is proven (Theorem \ref{T2}) from Corollary \ref{C1} and Lemma \ref{L1} by properties of the modal system. The properties required for the proof are the distribution principle and Brouwer's reduction theorem mentioned in Section \ref{sec:AxiomB}. The proof of Theorem \ref{T2} follows the proof of Sobel \citep{sobel2}[p. 126--127].

\setcounter{axiom}{0}
\setcounter{lemma}{0}
\setcounter{theorem}{0}
\setcounter{corollary}{0}
\setcounter{definition}{0}
